---
title: "IST 687 Final Project"
author: "Jonah Soos"
date: "2024-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in Relevant Packages
```{r}
library(tidyverse)
library(arrow)
library(lubridate)
library(stringr)
library(car)
library(mgcv)
library(caret)
library(Metrics)
library(xgboost)
library(fastDummies)
```

# Step 1 -- Initial Analysis of Data
```{r}
# Load in the Metadata
info <- read_csv("https://intro-datascience.s3.us-east-2.amazonaws.com/SC-data/data_dictionary.csv")

# Read in the Static House Information
house_data <- read_parquet("https://intro-datascience.s3.us-east-2.amazonaws.com/SC-data/static_house_info.parquet")
```

Firstly, because the data is extremely large, we are going to sample building IDs of 500 random houses to gather overall trends and observe summary statistics for the whole year. Typically we would use the entire dataset, but due to size constraints, sampling a proportion can allow us to get general trends without certainty.
```{r}
# Gather all unique Building IDS
building_ids <- unique(house_data$bldg_id)

# Creates blank data frame to hold data
energy_sample <- data.frame()

# Set seed for reproducibility
set.seed(11212024)

# Sample 25 different building_ids used for initial analysis
sample_buildings <- sample(building_ids, size = 600, replace = TRUE)

# Iterate through each unique building_id in the sample
for (i in sample_buildings) {
  # Create the unique URL for the building
  url <- paste0("https://intro-datascience.s3.us-east-2.amazonaws.com/SC-data/2023-houseData/", i, ".parquet")
  
  # Use TryCatch in case there are any timeout errors, the loop will not break
  tryCatch({
    # Read the parquet data
    temp <- read_parquet(url) %>%
      mutate(
        # Create a column that is the building id
        bldg_id = i,
        # Extract the Date
        date = as.Date(time),
        # Extract the Month,
        month = month(date),
        # Extract the Time
        timestamp = format(as.POSIXct(time), "%H:%M:%S"),
        # Calculate per hour by selecting all columns with "electricity" in the name and summing the row
        total_electricity = rowSums(select(., contains("electricity"))),
        # Calculate per hour by selecting all columns with "oil" in the name and summing the row
        total_oil = rowSums(select(., contains("oil"))),
        # Calculate per hour by selecting all columns with "gas" in the name and summing the row
        total_gas = rowSums(select(., contains("gas"))),
        # Calculate per hour by selecting all columns with "propane" in the name and summing the row
        total_propane = rowSums(select(., contains("propane"))),
        # Sum each factor to generate total energy usage
        total_energy_usage = total_electricity + total_gas + total_propane + total_oil
      )  
    
    # Bind the data to energy_data
    energy_sample <- rbind(energy_sample, temp)
    
    # Print progress of the loop
    print(paste0("Progress: ", which(sample_buildings == i) / length(sample_buildings) * 100, "%"))
  }, error = function(e) {
    # Handle the error (e.g., log it and continue)
    message(paste("Error with building ID:", i, "URL:", url, "Message:", e$message))
  })
}
```

Next, do some general data analysis for trends
```{r}

# Analyzing energy usage by month
monthly <- energy_sample %>%
  # Group the Data by month and house
  group_by(bldg_id, month) %>%
  # Exclude NAs for month
  filter(!is.na(month)) %>%
  # Calculate mean usage by month (optional if you want to include mean in the data)
  summarise(total_energy_usage = mean(total_energy_usage, na.rm = TRUE)) %>%
  # Create the boxplot with the data
  ggplot(mapping = aes(x = factor(month), y = total_energy_usage)) +
  geom_boxplot() +
  labs(
    title = "Monthly Energy Usage",
    x = "Month",
    y = "Total Energy Usage (kWh)"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 14)
  )
monthly
ggsave("monthly_energy.png", plot = monthly)

# Analyzing energy usage by hour
hourly <- energy_sample %>%
  # Group the Data by hour and house
  group_by(bldg_id, timestamp) %>%
  # Exclude NAs for hour
  filter(!is.na(timestamp)) %>%
  # Calculate mean usage by hour (optional if you want to include mean in the data)
  summarise(total_energy_usage = mean(total_energy_usage, na.rm = TRUE)) %>%
  # Create the boxplot with the data
  ggplot(mapping = aes(x = factor(timestamp), y = total_energy_usage)) +
  geom_boxplot() +
  labs(
    title = "Hourly Energy Usage",
    x = "Hour",
    y = "Total Energy Usage (kWh)"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 14)
  )
hourly
ggsave("hourly_energy.png", plot = hourly)

# Rank the highest and lowest contributors to energy consumption
ranked <- energy_sample %>%
  # summarise each energy contributor to find the mean
  summarise(
    across(out.electricity.ceiling_fan.energy_consumption:out.propane.range_oven.energy_consumption,
           ~ mean(., na.rm = TRUE))
  ) %>%
  # Use pivot longer to turn columns into rows
  pivot_longer(everything(), names_to = "Energy_Contributor", values_to = "Mean_Energy") %>%
  # Exclude anything with a mean of 0
  filter(Mean_Energy != 0) %>%
  # Edit the names so that it takes whatever text is after electricity|propane|oil and before energy_consuption
  mutate(
    Energy_Contributor = gsub(".*\\.(electricity|natural_gas|propane)\\.", "", Energy_Contributor),
    Energy_Contributor = gsub("\\.energy_consumption", "", Energy_Contributor)
  )

# Create the bar chart of the results
ggplot(ranked, aes(x = reorder(Energy_Contributor, Mean_Energy), y = Mean_Energy)) +
  geom_col(fill = "skyblue") +
  coord_flip() +
  labs(
    title = "Mean Energy Consumption by Contributor",
    x = "Energy Contributor",
    y = "Mean Energy Consumption per Hour (kWh)"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 14)
  ) -> vars
vars
ggsave("energy_impact.png", plot = vars)
```


Loading in all of the entire hourly energy data
```{r}
# Gather all unique Building IDS
building_ids <- unique(house_data$bldg_id)

# Creates blank data frame to hold data
energy_data <- data.frame()

# Iterate through each unique building_id
for (i in building_ids) {
  # Create the unique URL for the building
  url <- paste0("https://intro-datascience.s3.us-east-2.amazonaws.com/SC-data/2023-houseData/", i, ".parquet")
  
  # Use TryCatch in case there are any timeout errors, the loop will not break
  tryCatch({
    # Read the parquet data
    temp <- read_parquet(url) %>%
      mutate(
        # Create a column that is the building id
        bldg_id = i,
        # Extract the Date
        date = as.Date(time),
        # Extract the Month,
        month = month(date),
        # Extract the Time
        timestamp = format(as.POSIXct(time), "%H:%M:%S"),
        # Calculate per hour by selecting all columns with "electricity" in the name and summing the row
        total_electricity = rowSums(select(., contains("electricity"))),
        # Calculate per hour by selecting all columns with "oil" in the name and summing the row
        total_oil = rowSums(select(., contains("oil"))),
        # Calculate per hour by selecting all columns with "gas" in the name and summing the row
        total_gas = rowSums(select(., contains("gas"))),
        # Calculate per hour by selecting all columns with "propane" in the name and summing the row
        total_propane = rowSums(select(., contains("propane"))),
        # Sum each factor to generate total energy usage
        total_energy_usage = total_electricity + total_gas + total_propane + total_oil
      )  %>%
      # Filter to only summer months
      filter(month %in% (6:8)) %>%
      # Select relevant columns
      select(time:total_energy_usage)
    
    # Bind the data to energy_data
    energy_data <- rbind(energy_data, temp)
    
    # Print progress of the loop
    print(paste0("Progress: ", which(building_ids == i) / length(building_ids) * 100, "%"))
  }, error = function(e) {
    # Handle the error (e.g., log it and continue)
    message(paste("Error with building ID:", i, "URL:", url, "Message:", e$message))
  })
}
```


Loading in the Weather Data
```{r}
# Gather each unique county
counties <- unique(house_data$in.county)

# Create a dataframe for the weather data
weather_data <- data.frame()

# Iterate through each unique building_id
for (i in counties) {
  # Create the unique URL for the building
  url <- paste0("https://intro-datascience.s3.us-east-2.amazonaws.com/SC-data/weather/2023-weather-data/", i, ".csv")
  
  # Use TryCatch in case there are any timeout errors, the loop will not break
  tryCatch({
    # Read the csv data
    temp <- read_csv(url) %>%
      mutate(
        # Create a column that is the building id
        in.county = i,
        # Extract the Date
        date = as.Date(date_time),
        # Extract the Month,
        month = month(date_time),
        # Extract the Time
        timestamp = format(as.POSIXct(date_time), "%H:%M:%S")
      )  
    
    # Bind the data to energy_data
    weather_data <- rbind(weather_data, temp)
    
    # Print progress of the loop
    print(paste0("Progress: ", which(counties == i) / length(counties) * 100, "%"))
  }, error = function(e) {
    # Handle the error (e.g., log it and continue)
    message(paste("Error with building ID:", i, "URL:", url, "Message:", e$message))
  })
}

# Save the current environment so scraping does not need to occur multiple times
save.image("rawdata.RData")
```


#### Step 2 -- Creating Dataframe for Analysis

Merging all data together into a dataframe
```{r}

# Merging the data
data <- house_data %>%
  # Use left join to merge all matching building IDs onto house_data
  left_join(energy_data, by = c(
    "bldg_id" = "bldg_id"
  )) %>%
  # Use left_join to match county ids and timestamps with weather information
  left_join(weather_data, by = c(
    "in.county" = "in.county",
    "date" = "date",
    "month" = "month",
    "timestamp" = "timestamp"
  ))
write_parquet(data, "final_data.parquet")
```

If writing works properly, remove the raw data from the environment
```{r}
rm(energy_data, energy_sample, house_data, weather_data, temp)

# Save the current environment to load in final data if neccesary
save.image("rawdata.RData")
```

Use summary to make important takeaways
```{r}
summary(data)
```

Takeaways:
```{r}
# Upgrades and Weight are all the same, so do not include
# House are anywhere from 1:4 Bedrooms and 318:2176 square feet
# Location classifiers -- likely not needed because of weather data. Only 2 used (see below) where the difference can be quanitified by recorded humidity data
table(data$in.building_america_climate_zone)
table(data$in.census_region) # All from the same region, so no need to include
table(data$in.city) # 16 different classifications is likely too many to include, so will exclude
table(data$in.ceiling_fan) # 3 Different, may be interesting to look at fan usage
table(data$in.clothes_washer) # Differences but found not to be impactful earlier. Turn into dummy
table(data$in.clothes_dryer) # Differences but found not to be impactful earlier. Turn into dummy
table(data$in.cooking_range) # Differences but found not to be impactful earlier. Turn into dummy
table(data$in.cooling_setpoint) # Major differences that may impact energy, include as numeric
table(data$in.heating_setpoint) # Major differences that may impact energy, include as numeric
table(data$in.dishwasher) # Differences but found not to be impactful earlier. Turn into dummy
table(data$in.electric_vehicle) # No electric vehicles, do not include
table(data$in.geometry_stories) # Likely collinear with square feet, but include and check
table(data$in.heating_fuel) # Some differences that may impact energy consumption, include and check later
table(data$in.holiday_lighting) # No variation
table(data$in.hvac_cooling_efficiency) # Likely interesting to include, since there are a lot of classifiers, use cooling_type instead
table(data$in.hvac_heating_efficiency) # Likely interesting to include, since there are a lot of classifiers, use heating_type instead
table(data$in.lighting) # With lighting playing a strong factor, want to include
table(data$in.plug_loads) # All the same
table(data$in.vacancy_status) # Interesting, may impact energy usage
table(data$in.plug_loads) # Seen major factor previously, but zero variation, so not included
```



Edit the data to select important columns
```{r}

f_data <- data %>%
  # Select relevant columns
  select(
    bldg_id, in.county,
    # Important building parameters
    in.sqft, in.bedrooms, in.geometry_stories, in.occupants, in.lighting, in.vacancy_status,
    # Important appliances
    in.ceiling_fan, in.clothes_dryer, in.clothes_washer, in.cooking_range, in.dishwasher,
    # Cooling/heating types
    in.cooling_setpoint, in.heating_setpoint, in.hvac_cooling_type, in.hvac_heating_type, in.heating_fuel,
    # Demographic factors
    in.income, 
    # All energy columns
    date:total_energy_usage,
    # Weather columns
    temperature = `Dry Bulb Temperature [°C]`, 
    humidity = `Relative Humidity [%]`, 
    wind = `Wind Speed [m/s]`
  ) %>%
  # Apply data transformations
  mutate(
    # Make vacancy status a factor
    in.vacancy_status = as.factor(in.vacancy_status),
    # Turn appliances into categorical variables
    in.ceiling_fan = as.factor(in.ceiling_fan),
    # Turn lighting into categorical variable
    in.lighting = as.factor(in.lighting),
    # Turn occupants into numeric
    in.occupants = as.numeric(in.occupants),
    # Separate dryer type and efficiency
    dryer_type = as.factor(case_when(
      str_detect(in.clothes_dryer, "Electric") ~ "Electric",
      str_detect(in.clothes_dryer, "Gas") ~ "Gas",
      str_detect(in.clothes_dryer, "Propane") ~ "Propane",
      TRUE ~ "None"
    )),
    dryer_efficiency = case_when(
      str_detect(in.clothes_dryer, "120%") ~ 120,
      str_detect(in.clothes_dryer, "100%") ~ 100,
      str_detect(in.clothes_dryer, "80%") ~ 80,
      TRUE ~ NA_integer_
    ),
    # Separate washer presence and efficiency
    is_washer = case_when(
      str_detect(in.clothes_washer, "None") ~ 0,
      TRUE ~ 1
    ),
    washer_efficiency = case_when(
      str_detect(in.clothes_washer, "120%") ~ 120,
      str_detect(in.clothes_washer, "100%") ~ 100,
      str_detect(in.clothes_washer, "80%") ~ 80,
      TRUE ~ NA_integer_
    ),
    # Separate stove type and efficiency
    stove_type = as.factor(case_when(
      str_detect(in.cooking_range, "Electric") ~ "Electric",
      str_detect(in.cooking_range, "Gas") ~ "Gas",
      str_detect(in.cooking_range, "Propane") ~ "Propane",
      TRUE ~ "None"
    )),
    stove_efficiency = case_when(
      str_detect(in.cooking_range, "120%") ~ 120,
      str_detect(in.cooking_range, "100%") ~ 100,
      str_detect(in.cooking_range, "80%") ~ 80,
      TRUE ~ NA_integer_
    ),
    # Separate dishwasher presence and efficiency
    is_dishwasher = case_when(
      str_detect(in.dishwasher, "None") ~ 0,
      TRUE ~ 1
    ),
    dishwasher_efficiency = case_when(
      str_detect(in.dishwasher, "120%") ~ 120,
      str_detect(in.dishwasher, "100%") ~ 100,
      str_detect(in.dishwasher, "80%") ~ 80,
      TRUE ~ NA_integer_
    ),
    
    # Remove the "F" and extract numeric values for setpoints
    in.cooling_setpoint = as.numeric(str_extract(in.cooling_setpoint, "\\d{2}")),
    in.heating_setpoint = as.numeric(str_extract(in.heating_setpoint, "\\d{2}")),
    # Convert cooling/heating types and heating fuel to factors
    in.hvac_cooling_type = as.factor(in.hvac_cooling_type),
    in.hvac_heating_type = as.factor(in.hvac_heating_type),
    in.heating_fuel = as.factor(in.heating_fuel)
  ) %>%
  # Drop unnecessary columns
  select(-c(in.clothes_dryer, in.clothes_washer, in.dishwasher, in.cooking_range))

# Remove the older version for storage purposes
rm(data)

save.image("finaldata.RData")
```

#### Step 3 -- Modeling

```{r}
# First step is removing all identifying variables and variables used to create total_energy_usage

mod_data <- f_data %>%
  select(-c(bldg_id, in.county, date, total_electricity:total_propane)) %>%
  mutate(
    in.income = as.factor(in.income),
    timestamp = as.factor(timestamp)
  )

# Next create an OLS model as the baseline

# Perform train-test split
set.seed(1232024)  # For reproducibility
train_index <- createDataPartition(mod_data$total_energy_usage, p = 0.8, list = FALSE)

# Create training and testing datasets
train <- mod_data[train_index, ]
test <- mod_data[-train_index, ]

# Run the model with all variables
ols_model <- lm(total_energy_usage ~ ., data = train)
summary(ols_model)

# Takeaways -- Model does not fully work because of some factors being ordered wrong and efficiency being highly colinear between appliances. That can be seen here

# Filter to numeric data and create a corelation matrix
numeric_data <- mod_data %>% select(where(is.numeric))
cor_matrix <- cor(numeric_data, use = "pairwise.complete.obs")

# Get the correlations with magnitude > 0.5
high_corr <- as.data.frame(as.table(cor_matrix)) %>% 
  filter(abs(Freq) > 0.5 & Var1 != Var2)  # Exclude self-correlations

# Sort by descending correlation magnitude
high_corr <- high_corr %>% arrange(desc(abs(Freq)))

# Print the results
print(high_corr)
rm(numeric_data)

# Knowing this, we can include only an overall efficiency metric and reorder some of the factors perviously (so that none is the excluded factor)

mod_data <- mod_data %>%
  mutate(
    # Create appliance efficiency (can equal any specific one since all are equal)
    appliance_efficiency = dryer_efficiency,
    # Reordering certain factors
    in.hvac_heating_type = relevel(mod_data$in.hvac_heating_type, ref = "None"),
    in.heating_fuel = relevel(mod_data$in.heating_fuel, ref = "None"),
    in.income = relevel(mod_data$in.income, ref = "<10000"),
    stove_type = relevel(mod_data$stove_type, ref = "None"),
  ) %>%
  select(-c(washer_efficiency, dryer_efficiency, dishwasher_efficiency, stove_efficiency))

# Now with this dataset, we can rerun the OLS model
train <- mod_data[train_index, ]
test <- mod_data[-train_index, ]

# Run the model with all variables
ols_model <- lm(total_energy_usage ~ ., data = train)
summary(ols_model)

# Now, It can be seen heating type and heating fuel are collinear as well as there being a washer and dryer. So we can drop is_washer and see which of heating fuel and hvac heating has a larger impact on the model
mod_data <- mod_data %>% select(-is_washer)

# Find impact of in.hvac_heating_type
hvac <- lm(total_energy_usage ~ in.hvac_heating_type, mod_data)
summary(hvac)
# Find impact of in.heating_fuel
fuel <- lm(total_energy_usage ~ in.heating_fuel, mod_data)
summary(fuel)

# Both contain sigularities and both have minimal impact, so choosing to not include either
mod_data <- mod_data %>% select(-in.heating_fuel, -in.hvac_heating_type,
                                # Also removing income as a variable as its nature of factor makes                                             conclusions nearly impossible to disentangle
                                -in.income)

# Resplit the data
train <- mod_data[train_index, ]
test <- mod_data[-train_index, ]

# Fix Error in Dryer Type
test$dryer_type <- ifelse(test$dryer_type == "None", "Unknown", as.character(test$dryer_type))
test$dryer_type <- factor(test$dryer_type, levels = levels(train$dryer_type))

# Run the model with all variables
ols_model <- lm(total_energy_usage ~ ., data = train)
summary(ols_model)
vif(ols_model) # Check for multicollinearity

# Predict using the model to compute the RMSE
test$predicted <- predict(ols_model, newdata = test)

# Filter out NAs
test <- test %>%
  filter(!is.na(total_energy_usage & predicted))
ols_RMSE <- Metrics::rmse(test$total_energy_usage, test$predicted)
print(ols_RMSE)

save.image("mod.RData")
```

The base RMSE we can try and top is 0.5913. Lets first try dropping the linearity assumption using a GAM model to look for smooth relationships between regressors and independent variable

```{r}
# Fit a GAM model with smooth terms for continuous predictors with custom degrees of freedom to make sure there are not more than the number of different values
sapply(train, function(x) length(unique(x)))

gam_model <- gam(total_energy_usage ~ 
                 s(in.sqft, k = 8) + s(in.bedrooms, k = 4) + s(in.geometry_stories, k = 2) + 
                 s(in.occupants, k = 9) + factor(in.lighting) + 
                 factor(in.vacancy_status) + factor(in.ceiling_fan) +
                 s(in.cooling_setpoint, k = 10) + s(in.heating_setpoint, k = 11) + 
                 factor(in.hvac_cooling_type) + month + factor(timestamp) + 
                 s(temperature) + s(humidity) + s(wind) + 
                 factor(dryer_type) + factor(stove_type) + is_dishwasher +
                 s(appliance_efficiency, k = 3), 
                 data = train)

# Summary of the GAM model
summary(gam_model)

# Predict using the GAM model
test$predicted_gam <- predict(gam_model, newdata = test)

# Calculate RMSE (ensure there are no NA values in predictions)
rmse_gam <- rmse(test$total_energy_usage[!is.na(test$total_energy_usage)],
                   test$predicted_gam[!is.na(test$predicted_gam)])

# Print RMSE
print(rmse_gam)
save.image("mod.RData")

```

```{r}
plot(gam_model, pages = 10, se = TRUE, rug = TRUE, ylab = "Impact on Energy Usage", ylim = c(-2, 2))
```

Test a XGBoost model for Machine Learning

```{r}
# Turn all categorical variables to dummies for machine learning
train_processed <- train %>%
  dummy_cols(remove_first_dummy = TRUE, remove_selected_columns = TRUE)

test_processed <- test %>%
  dummy_cols(remove_first_dummy = TRUE, remove_selected_columns = TRUE)

# Keep only the columns in test_processed that are also in train_processed
common_columns <- intersect(names(train_processed), names(test_processed))
test_processed <- test_processed[, common_columns]

# Create xgb matrices for the data, used to train and test using the xgboost package
dtrain <- xgb.DMatrix(as.matrix(select(train_processed, -total_energy_usage)),
                          label = train$total_energy_usage)
dtest <- xgb.DMatrix(as.matrix(select(test_processed, -total_energy_usage)),
                          label = test$total_energy_usage)

rm(train_processed, test_processed)

# Set the parameters to default
params <- list(
  objective = "reg:squarederror",  # For regression
  eval_metric = "rmse",           # Root Mean Squared Error
  eta = 0.05                       # Learning rate
)

# Run the XGBoost Model
xgb_model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 100,          # Number of boosting rounds
  watchlist = list(train = dtrain, test = dtest),  # Monitor training and testing
  verbose = 1             # Show training progress
)

xgb_rmse <- min(xgb_model$evaluation_log$test_rmse)
```

The XGBoost model comes in similar RMSE, if not slightly better, but loses the intuitiveness component of understandable statistically backed takeaways. For that reason, the marginal benefit is not worth the loss of intuitiveness.

#### Step 4 -- Peak Energy Demand

```{r}
# Create a dataset for the month of july with temperature 5 degrees warmer
rm(train, test)

future_energy <- f_data %>%
  # Filter to July
  filter(month == 7) %>%
  # Make all Temps 5 degrees warmer
  mutate(temperature = temperature + 5) %>%
  # Remove the previous energy usage
  select(-total_energy_usage) %>%
  # Make the changes above
  mutate(
    # Create appliance efficiency (can equal any specific one since all are equal)
    appliance_efficiency = dryer_efficiency,
    timestamp = as.factor(timestamp)
  ) 

# Predict Energy for the new dataset
future_energy$total_energy_usage <- predict(gam_model, newdata = future_energy)
write_parquet(future_energy, "future_energy.parquet")

# Aggregate by day to find peak daily demand
future_hourly_demand <- future_energy %>%
  group_by(timestamp) %>%
  summarise(future_energy_usage = sum(total_energy_usage, na.rm = TRUE))

# Aggregate current demand usage to compare
current_hourly_demand <- f_data %>%
  filter(month == 7) %>%
  group_by(timestamp) %>%
  summarise(current_energy_usage = sum(total_energy_usage, na.rm = TRUE))

# Join data sets and calculate the percent change per hour
demand <- current_hourly_demand %>%
  left_join(future_hourly_demand, by=c("timestamp")) %>%
  mutate(
    pct_change = (future_energy_usage - current_energy_usage) / current_energy_usage
  )

# Reshape the data to long format
demand_long <- demand %>%
  pivot_longer(cols = c(current_energy_usage, future_energy_usage),
               names_to = "energy_type",
               values_to = "energy_usage")

# Create the bar graph with side-by-side bars
ggplot(demand_long, aes(x = timestamp, y = energy_usage, fill = energy_type)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.4) +  # Side-by-side bars
  scale_fill_manual(
    name = "Energy Type",
    values = c("current_energy_usage" = "blue", "future_energy_usage" = "red")
  ) +
  labs(
    title = "Current vs. Future Energy Usage by Time",
    x = "Timestamp",
    y = "Energy Usage",
    fill = "Energy Type"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 14))

```

Evaluating Peak Energy Demand by different variables

```{r}
# Demand by County

# Evaluate Future County Demand
cf_demand <- future_energy %>%
  group_by(in.county) %>%
  summarise(future_energy_usage = sum(total_energy_usage, na.rm = TRUE))

# Evaluate Current County Demand
cc_demand <- f_data %>%
  filter(month == 7) %>%
  group_by(in.county) %>%
  summarise(current_energy_usage = sum(total_energy_usage, na.rm = TRUE))

# Join data sets
c_demand <- cc_demand %>%
  left_join(cf_demand, by=c("in.county"))

# Reshape the data to long format
c_demand_long <- c_demand %>%
  pivot_longer(cols = c(current_energy_usage, future_energy_usage),
               names_to = "energy_type",
               values_to = "energy_usage")

# Create the bar graph with side-by-side bars
ggplot(c_demand_long, aes(x = in.county, y = energy_usage, fill = energy_type)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.4) +  # Side-by-side bars
  scale_fill_manual(
    name = "Energy Type",
    values = c("current_energy_usage" = "blue", "future_energy_usage" = "red")
  ) +
  labs(
    title = "Current vs. Future Energy Usage by County",
    x = "Timestamp",
    y = "Energy Usage",
    fill = "Energy Type"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 14))

# Demand by Occupants

# Evaluate Future County Demand
of_demand <- future_energy %>%
  group_by(in.occupants) %>%
  summarise(future_energy_usage = sum(total_energy_usage, na.rm = TRUE))

# Evaluate Current County Demand
oc_demand <- f_data %>%
  filter(month == 7) %>%
  group_by(in.occupants) %>%
  summarise(current_energy_usage = sum(total_energy_usage, na.rm = TRUE))

# Join data sets
o_demand <- oc_demand %>%
  left_join(of_demand, by=c("in.occupants"))

# Reshape the data to long format
o_demand_long <- o_demand %>%
  pivot_longer(cols = c(current_energy_usage, future_energy_usage),
               names_to = "energy_type",
               values_to = "energy_usage")

# Create the bar graph with side-by-side bars
ggplot(o_demand_long, aes(x = in.occupants, y = energy_usage, fill = energy_type)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.4) +  # Side-by-side bars
  scale_fill_manual(
    name = "Energy Type",
    values = c("current_energy_usage" = "blue", "future_energy_usage" = "red")
  ) +
  labs(
    title = "Current vs. Future Energy Usage by Occupants",
    x = "Occupants",
    y = "Energy Usage",
    fill = "Energy Type"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 14))
```

Stargazer output of GAM Final Model
```{r}
library(mgcv)
library(stargazer)
# Generate summary statistics from the GAM model
gam_summary <- summary(gam_model)

# Extract coefficients and standard errors
coefficients <- gam_summary$p.table
smooth_terms <- gam_summary$s.table

# Create a table with parametric terms
stargazer(coefficients, 
          type = "html", 
          summary = FALSE, 
          title = "Parametric Coefficients for GAM Model",
          out = "gam_parametric_coefficients.html")

# Optional: Create a separate table for smooth terms
cat("\nSmooth Terms:\n")
stargazer(smooth_terms, 
          type = "html", 
          summary = FALSE, 
          title = "Smooth Terms for GAM Model",
          out = "gam_smooth_terms.html")
```

